---
title: "PA1_template"
author: "Federico Roscioli"
date: "5/13/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible Research: Peer Assessment Week 2

## Load and process the data
### 1. Load the data
```{r}
Sys.setlocale("LC_TIME", "en_US")
library(dplyr)
library(ggplot2)
library(lubridate)
fileurl <-
        "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zipfile <- "./course5week2/data.zip"
# for downloading the file as data.zip to the folder course4week4 under the working
# directory
filedir <- "./course5week2"
unzip_path <-
        "./course5week2/data"  ##### path for storing the unzipped files #######
if (!file.exists(filedir)) {
        dir.create(filedir)
}
download.file(fileurl, file.path(zipfile))
unzip(zipfile, exdir = unzip_path) ####### exdir is the extract directory ##########
datafile <- file.path(unzip_path, "activity.csv")

activity <- read.csv(datafile)
```

### 2. Fixing date var and adding weekdays
```{r}
activity$date <- ymd(activity$date)
activity$weekend <-
        as.factor(ifelse(
                weekdays(activity$date) == "Saturday" |
                        weekdays(activity$date) == "Sunday",
                "weekend",
                "weekday"
        ))
activity$dayofweek <- as.factor(weekdays(activity$date))
```

### 3. Histogram of the steps per day
```{r}
stepsByDay <-
        activity %>% group_by(date) %>% summarise(stepsperday = sum(steps, na.rm = TRUE))
qplot(
        stepsperday,
        data = stepsByDay,
        na.rm = TRUE,
        binwidth = 500,
        xlab = 'Total steps per day',
        ylab = 'Frequency using binwith 500',
        main = 'Histogram of the total number of steps taken each day'
)
```

### 4. Mean total number of steps taken per day
```{r}
meanstepsperday <-
        stepsByDay %>% summarise(
                average = mean(stepsperday, na.rm = TRUE),
                median = median(stepsperday, na.rm = TRUE)
        )
meanstepsperday
```

### 5. Avarage number of steps per time interval (5 min)
```{r}
interval_average <-
        activity %>% group_by(interval) %>% summarise(average = mean(steps, na.rm = TRUE))
qplot(
        interval,
        average,
        data = interval_average,
        geom = "line",
        xlab = "5-minute intervals",
        ylab = "Average steps taken across all days"
)
```

### 6. Max steps interval during the day
```{r}
maxint <- interval_average[which.max(interval_average$average), ]
maxint$interval
```

### 7. Number of NA
```{r}
NAnumb <- sum(is.na(activity))
NAnumb
```

### 8. Substituting NA
The choice is to substitute the NA with the avarage number of steps for eache interval
```{r}
# subset dataset where there are no NAs
activity_no_NA <- activity[which(!is.na(activity$steps)), ]

# calculate the mean steps for each interval
interval_only <-
        activity_no_NA %>% group_by(interval) %>% summarise(average = mean(steps))

# convert the average to integer
interval_only$average <- as.integer(interval_only$average)

#subset dataset where steps have NAs
activity_na <- activity[which(is.na(activity$steps)), ]

# fill NAs with average steps based on interval
activity_na$steps <-
        ifelse(activity_na$interval == interval_only$interval,
               interval_only$average)

# row bind the datasets that do not have NAs and the dataset where NAs are replaced with
# mean values
activity_impute <- rbind(activity_no_NA, activity_na)
```

### 9. Hitogram after NA substitution
```{r}
stepsByDay_impute <-
        activity_impute %>% group_by(date) %>% summarise(stepsperday = sum(steps))
qplot(
        stepsperday,
        data = stepsByDay_impute,
        na.rm = TRUE,
        binwidth = 500,
        xlab = 'Total steps per day',
        ylab = 'Frequency using binwith 500',
        main = 'Histogram of the total number of steps taken each day'
)
```

### 10. Mean and median after NA substitution
```{r}
totalstepsperday_impute <-
        activity_impute %>% group_by(date) %>% summarise(stepsperday = sum(steps))
mean_n_median <-
        totalstepsperday_impute %>% summarise(average = mean(stepsperday),
                                              median = median(stepsperday))
mean_n_median
```

### 11. Ddifference between weekends and weekdays
```{r}
meansteps <-
        activity_impute %>% group_by(interval, weekend) %>%   summarise(average = mean(steps))
qplot(
        interval,
        average,
        data = meansteps,
        geom = "line",
        xlab = "5-minute interval",
        ylab = "average number of steps",
        main = "Average steps pattern between Weekday and Weekend",
        facets = weekend ~ .
)
```



